OpenMP is Not as Easy as It Appears, \\ but It is the Easiest One

### 2017-07-13 sections/01-introduction.tex
Re-organizing and fixing the used template

There is an ongoing evolution of parallel processing platforms. The pursuit for high performance in general-purpose and scientific applications have turned to the exploration of the parallel processing capabilities provided by multi-core processors and arrangements of coprocessors like the Xeon Phi~\cite{xeonphiRuntime}. The machine currently at the top of the \textit{TOP500} list~\cite{dongarra1997top500} uses multi-core processors (12 cores) and Xeon Phi coprocessors~\cite{top500:site}.
Multi-core processors have become popular on personal computers and mobile devices. This popularization increases the need for specialized developers able to work in parallel applications and apply parallel computing models~\cite{Chen:2009:ChallengesMIC}~\cite{okur2012developers}.

Today there is a wide gap between legacy code applications that could benefit from parallel computing power and the new parallel computing platforms that could provide that power. However, legacy applications are generally sequential and therefore not easily scalable to multi-core architectures and parallel computing platforms.

Asanovic \textit{et al.} \cite{Asanovic2009} describes this scenario, and defends the need for Structural and Computational Patterns in Parallel Computing. These patterns allow the portability of parallel programming solutions, while also allowing specialization of applications with the same behavior. In some situations it is not a matter of achieving the highest possible performance, but instead of making the application achieve acceptable performance, benefiting from a new parallel computing platform~\cite{Asanovic2009}. A similar approach is proposed by Pllana \textit{et al.} in a development environment that uses a combination of Model-Driven Development and Parallel Building Blocks \cite{Pllana:2009}. The approach consists in creating the model and then generating or transforming the code using \emph{model-to-source} tools.

@@

We argue that it is necessary to consider fundamental concepts of architectures and parallel programming techniques when teaching students new to the subject, so that they can learn when and how to apply the programming models~\cite{ben1999thinking}. The abstractions provided by parallel programming tools over concepts such as threads, barriers and loop parallelization should be carefully studied, analysed and conceptually built in the classroom before they are used by the students in their assignments.

We found eight tutorials with different kinds of \texttt{OpenMP} errors in the experiment with the students' assignments. Many of these errors where similar to the most common \texttt{OpenMP} mistakes~\cite{SuB:2005:CMO:1892830.1892863}. The most common type of error identified in our experiment was the management of shared or private variables. This problem is common across multiple programming paradigms.

The \textbf{main contribution of this work} is to show the importance of teaching core architecture and parallel programming concepts thoroughly, despite the availability of powerful tools such as \texttt{OpenMP}. We also aim to show that the \texttt{OpenMP} API is not as easy to use as it might seem.

The rest of the paper is organized as follows. The following section discusses related work. Section~\ref{sec:concurrency} discusses concurrency and other concepts in parallel computing. Section~\ref{sec:directives} presents the idea of compilation directives. Section~\ref{sec:methodology} discusses the methodology applied in the experiments, and Section~\ref{sec:results} presents our results. Section~\ref{sec:students:perception} presents the perceptions of the graduate students regarding the exercise and learning \texttt{OpenMP}. Finally, Section~\ref{sec:conclusions} presents the conclusions and next steps.

% Institution 2 = University of São Paulo

>>>>

...

@@

The rest of the paper is organized as follows. The following section discusses
related work...

### 2017-07-13 sections/03-relatedwork.tex
Re-organizing and fixing the used template

There has been a large body of works that have presented results of studies and experiments in teaching parallel computing courses. S\"{u}\ss~and Leopold~\cite{SuB:2005:CMO:1892830.1892863} discuss the teaching of parallel computing, in particular the use of \texttt{OpenMP}. The authors present the most common errors identified in assignments given in parallel computing classes. The errors were classified in two categories: \emph{correctness mistakes} and \emph{performance mistakes}. Errors that impact the correctness of a program are errors of the first category. Errors which make programs perform slower are classified in the second category. The work also provides a checklist containing tips to avoid the most common mistakes.

Falcao~\cite{6565518} defends that teaching a parallel programming model or interface such as \texttt{OpenMP} in the beginning of undergraduate courses is not much harder than teaching a language like C or C++. The benefits of teaching these skills early on in the course would be interesting, considering that the majority of computer architectures are now are multi-core. Parallel computing should be taught with the use of higher level tools. One of the approaches proposed by Ferner \textit{et al.}~\cite{6651019} is to use compiler directives to describe how a program should be parallelized.

The need for Structural and Computational Patterns to Parallel Computing is defended by Asanovic \textit{et al.}~\cite{Asanovic2009} as the way to allow portability of solutions and the specialization of applications with the same behavior. Using patterns to design applications make it easy to change the target platform. The applications will be bound to platforms in low levels of architecture and probably inside a specialized function which can be easily changed.

A similar approach is proposed by Pllana \textit{et al.}~\cite{Pllana:2009} in a programming environment that uses a combination of Model-Driven Development and Parallel Building Blocks. The idea is to create the model and then generate or transform the code using model-to-source tools. The proposed environment uses intelligent agents to assist the programmer, this is done during the high level program composition and when the programmer loads a BLAS function for some dense linear algebra operation, the tool suggests the use of the appropriate parallel basic block. The parallelism is inside the parallel building blocks and the programmer is not exposed directly to the complexity of parallel programming problems.

Speyer \textit{et al.}~\cite{4755913} defend that implementing applications with high performance is harder than ever, specially due to the emergence of multi-core architectures. This leads to the development of hybrid applications that use both the \texttt{OpenMP} and \texttt{MPI} (\emph{Message Passing Interface}) APIs. This presents even greater challenges to the programmers, since they now need to know how to use both code annotations and message passing abstractions to achieve high performance. The paper then compares different parallel programming tools with regard to their ease of use. There is a substantial effort to start writing parallel and high performance code even with these tools.

Some problems can be solved using different directive combinations, and the chosen combination influences the code that is generated. The \texttt{Parallware}~\cite{Parallware:2014} source-to-source compiler showed how different strategies and uses of directive combinations can lead to different performance results. The use of the wrong \texttt{OpenMP} directives and strategies can significantly reduce the performance over a sequential version.

The \texttt{OmpVerify}~\cite{Basupalli:2011} system implements a static analysis tool based on the polyhedral model to extract the parallel portions of a program automatically and to detect important classes of common data-race errors in \texttt{OpenMP} \texttt{parallel for} directives. Saillard \textit{et al.}~\cite{Saillard:2014} statically validates barriers and work sharing constructs to avoid the improper use of directives that could cause deadlocks. S\"{u}\ss~and Leopold~\cite{Leopold:userOpenMP}, present an user's experience with parallel sorting and OpenMP. They have used several versions of a simple parallel sorting algorithm to show some weaknesses of OpenMP.

Substantial efforts have been made to achieve the simplest possible way to parallelize code, and the task is not easy. In this paper we present an exploratory analysis of common \texttt{OpenMP} errors, confirming that programming with high-level APIs and abstractions can still be difficult.

>>>>

TO-DO ...
